Cosmology Essay: A review of Dark Matter and Dark Energy

### 2017-05-05 cosmology/Essay/CosmoEssayHK.tex
cosmology

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.jpg}
\author{Howard Kinsman}
\title{Cosmology Essay: A review of Dark Matter and Dark Energy}
\begin{document}
\maketitle
\section{Introduction}
Many observations, and subequent cosmological theories, have found that approximately 95\% of the mass and energy in the universe is dark and invisible, the true nature of which is
largely unknown. The matter component is known as dark matter (DM) and the energy component dark energy (DE) This is a brief review of the history and current status of 
DM and DE.
\section{Dark Matter}
DM has a surprisingly long and interesting history. In 1933, the renowned astrophysicist, Fritz Zwicky published a paper on the velocity dispersion of galaxies within the Coma cluster. He found
that the velocity dispersion of galaxies within the cluster was so high that in order for the system to remain stable there must be some ``missing mass'' i.e. mass which
is not detectable optically. Further research in this area continued but it wasn't until the 1970s when astronomers began accurately measuring the rotation curves of spiral galaxies,
with the discovery of HI 21cm radio emission, that interest in this area became of great importance. HI 21cm radio emission is produced by neutral hydrogen during a change of energy
state. It has a very precise wavelength of 21cm.
Rotation curves are a measure of the orbital velocity of the stars and gas in a galaxy plotted as a function of distance from the galactic centre. According to
Keplerian laws rotation curves would be expected to decline with increasing radius however it was found that they remain flat i.e. constant, all the way out to the largest radii observed, 
both optically and with 21 cm. According to Newtonian gravity this implies that there must be some ``hidden mass'' within the galaxies exerting a gravitational effect on the system resulting
in these flat rotation curves. This ``hidden mass'' was termed dark matter.
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./rotationcurves.jpg}
\caption{Examples of flat galaxy rotation curves \citep{ruben}}
\label{fig:1}
\end{figure}

Apart from high velocity dispersions in galaxy clusters and flat rotation curves other evidence for DM includes gravitational lensing, hot gas in clusters, the Cosmic Microwave Background (CMB)
 and through computer modelling of the growth of 
large scale structure in the early universe (discussed later). Gravitational lensing occurs when light from distant galaxies behind a galaxy cluster is distorted due to the 
large gravitational pull of the cluster. This is known as strong lensing and is quite rare, however weak lensing also occurs where there is just a small distortion of the galaxies size
and shape. 
The mass of the galaxy cluster can be measured and compared to how much the light is expected to be bent by gravity. The results show that there must be far more mass in clusters 
than can be detected. Galaxy clusters when viewed in X-ray reveal huge amounts of hot gas which can only be explained by large amounts of DM which provide a potential gravitational well to hold the gas.
The CMB is the remnants of radiation from the early universe. Anisotropies in the CMB also reveal evidence for DM. Anisotropies are irregularities seen in the CMB which, after having been
thoroughly studied, reveal what woud be expected if thermal variations in a small space had grown to form the observable universe today.
Before decoupling from baryonic matter (i.e. ordinary matter comprised of 
mainly protons and neutrons), photons underwent oscillations that froze in at redshift $z\approx1100$. Decoupling refers to events in the early universe when different
particles fall out of thermal equilibrium with each other.
By studying the height of the peaks in these oscillations it has been 
found that the universe is comprised of approximately 5\% baryonic matter, 26\% dark matter
and the remainder dark energy (discussed later). Some authors \citep{freese} regard the evidence of DM from the CMB as ``irrefutable''.

@@

With the advent of numerical simulations, models that involve DM have been extensively explored. Currently a very popular model of the evolution of structure in the universe is
the Lambda Cold Dark Matter ($\Lambda CDM$) model. $\Lambda$ here refers to the cosmological contant discussed in the DE section. 
Computer simulations have shown that if the DM particles are relativistic (``hot'') then, on small scales, density fluctuations are Silk damped, or
washed out, the random thermal motion of DM particles. This results in small scale structure fluctuations to be suppressed. However, if the DM particles are ``cold'' i.e. cold dark matter (CDM), then
they undergo a very different structure formation because they have a much smaller free-streaming length and can form low mass halos which can gradually build up into larger DM
structures. These DM structures provide a gravitational well that baryonic particles can fall into to form baryonic structure. This results in a bottom-up process of structure 
formation which is very different to the top-down sequence predicted for hot DM. These simulations have further convinced cosmologists that CDM comprised of WIMPs are the best
explanation for the growth of large scale structure in the universe. In fact, the $\Lambda CDM$ model agrees so well with observations that it is often regarded as the ``standard model
of cosmology''.

@@

Type Ia supernovae occur when a white dwarf in a binary pair accretes enough mass so that it goes over the Chandrasekhar mass and triggers a supernova. The supernova therefore doesn't
depend on the mass of the white dwarf on the other star in the binary. Type Ia supernovae can be considered as standard candles and enable astronomers to obtain a luminosity distance.
To measure cosmic expansion the apparent magnitude of distant supernovae are compared to those of more local supernovae which are not affected by DE. The Supernova Cosmology Project
which carried out one of the surveys in 1998 found that, with a flat cosmology ($\Omega_{\Lambda}+\Omega_m=1$), that $\Omega_m=.28$ and so that $\Omega_{\Lambda}=.72$. $\Omega$ is the
density parameter and is defined as the ratio of the observed density to the critical density. $\Omega_m$ here is 
the proportion of matter in universe (DM and baryonic) whilst $\Omega_{\Lambda}$ represents the proportion of DE in universe. The deceleration parameter is calculated as follows:
\begin{equation}
\frac{1}{2}(\Omega_m-2\Omega_{\Lambda}+2\Omega_R) 
\end{equation}
This then results in a negative deceleration parameter indicating that the expansion of the universe is accelerating.
Since 1998 there have been many more supernova surveys and it is has become an active area of observational cosmology. All this supernova data indicates that the apparent luminosity
of Type Ia supernovae declines more rapidly with increasing redshift than would be expected if $\Omega_M=1$ and $\Omega_{\Lambda}=0$ i.e. an Einstein-de Sitter universe, or the 
luminosity distance is greater than expected. Of course, this data could be interpreted as being due to other effects such as dust or gas in the line of sight. More evidence is required.

@@

There are many theories for DE. However, the simplest one is that it is constant i.e. a cosmological constant. The cosmological constant has a negative pressure equal to
its energy density and so exerts a repulsive force causing the expansion of the universe to accelerate i.e. $P=-\rho$. The previously mentioned $\Lambda CDM$ model uses
this form of DE and is the currently preferred cosmological model. An important parameter in the study of DE is the equation of state parameter $\omega$. This comes from: 
\begin{equation}
P=\omega\rho c^2
\end{equation}
where P is pressue, $\rho$ is density, c is speed of light and $\omega$ is the equation of state parameter.
In the $\Lambda CDM$ model $\omega$ is assumed to be -1.

@@

A new and exciting development by the European Space Agency, which will further probe DE, will be the launch of Euclid which is expected to be in 2020. Euclid aims to find out
the nature the acceleration of the expansion of the universe and DE. It will measure shapes and redshifts of galaxies and also explore the large scale structure of the universe
by measuring the distribution of galaxy clusters. Euclid is a satellite equipped with a 1.2m telescope and three imaging and spectroscopic instruments working in optical and
near-infrared. It will be placed at the L2 Langrange point, a stable point in space, and is planned to operate for six years. The plan is cover 15000 square degrees of sky and
image one billion galaxies and take redshifts measurements of 100,000 galaxies \citep{amendola}. With this data Euclid will able to detail clustering of galaxies out to redshift
z=2 and out to redshift z=3 utilising weak lensing. Redshift measurements and weak lensing will also accompanied by measuring correlations with the CMB, obtaining galaxy cluster 
information, strong lensing and possibly even obtaining luminosity distance through supernovae Type Ia measurements.

DE may well have a profound effect on the final fate of the universe. Observations suggest that the expansion of the universe will continue to accelerate forever and so the 
universe will cool as it expands. Redshift will eventually stretch photons to undetectable wavelengths and galaxies will therefore appear to disappear. Star formation will cease 
as gas supply runs out. Stars will
eventually die out, one by one, leaving stellar remnants. Some theories even predict that stellar remnants will disappear due to proton decay leaving only black holes,
which themselves will eventually disappear due to Hawking radiation. Proton decay is a hypothetical form of radiactive decay in which protons decay into ligher particles. 
Hawking radiation is blackbody radiation believed to be emitted by black holes because of quantum
effects near the event horizon. This is often referred to as the heat death of the universe. It would seem that the ultimate fate of our universe could potentially be a very 
cold, dark void.

>>>>

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.jpg}
\author{Howard Kinsman}
\title{Cosmology Essay: A review of Dark Matter and Dark Energy}
\begin{document}
\maketitle
\section{Introduction}
Many observations, and subsequent cosmological theories, have found that approximately 95\% of the mass and energy in the universe is dark and invisible, the true nature of which is
largely unknown. The matter component is known as dark matter (DM) and the energy component dark energy (DE) This is a brief review of the history and current status of 
DM and DE.
\section{Dark Matter}
DM has a surprisingly long and interesting history. In 1933, the renowned astrophysicist, Fritz Zwicky published a paper on the velocity dispersion of galaxies within the Coma cluster. He found
that the velocity dispersion of galaxies within the cluster was so high that in order for the system to remain stable there must be some ``missing mass'' i.e. mass which
is not detectable optically. Further research in this area continued but it wasn't until the 1970s when astronomers began accurately measuring the rotation curves of spiral galaxies,
with the discovery of HI 21cm radio emission, that interest in this area became of great importance. HI 21cm radio emission is produced by neutral hydrogen during a change of energy
state. It has a very precise wavelength of 21cm.
Rotation curves are a measure of the orbital velocity of the stars and gas in a galaxy plotted as a function of distance from the galactic centre. According to
Keplerian laws rotation curves would be expected to decline with increasing radius however it was found that they remain flat i.e. constant, all the way out to the largest radii observed, 
both optically and with 21 cm. According to Newtonian gravity this implies that there must be some ``hidden mass'' within the galaxies exerting a gravitational effect on the system resulting
in these flat rotation curves. This ``hidden mass'' was termed dark matter.
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./rotationcurves.jpg}
\caption{Examples of flat galaxy rotation curves \citep{ruben}}
\label{fig:1}
\end{figure}

Apart from high velocity dispersions in galaxy clusters and flat rotation curves other evidence for DM includes gravitational lensing, hot gas in clusters, the Cosmic Microwave Background (CMB)
 and through computer modelling of the growth of 
large scale structure in the early universe (discussed later). Gravitational lensing occurs when light from distant galaxies behind a galaxy cluster is distorted due to the 
large gravitational pull of the cluster. This is known as strong lensing and is quite rare, however weak lensing also occurs where there is just a small distortion of the galaxies size
and shape. 
The mass of the galaxy cluster can be measured and compared to how much the light is expected to be bent by gravity. The results show that there must be far more mass in clusters 
than can be detected. Galaxy clusters when viewed in X-ray reveal huge amounts of hot gas which can only be explained by large amounts of DM which provide a potential gravitational well to hold the gas.
The CMB is the remnants of radiation from the early universe. Anisotropies in the CMB also reveal evidence for DM. Anisotropies are irregularities seen in the CMB which, after having been
thoroughly studied, reveal what would be expected if thermal variations in a small space had grown to form the observable universe today.
Before decoupling from baryonic matter (i.e. ordinary matter comprised of 
mainly protons and neutrons), photons underwent oscillations that froze in at redshift $z\approx1100$. Decoupling refers to events in the early universe when different
particles fall out of thermal equilibrium with each other.
By studying the height of the peaks in these oscillations it has been 
found that the universe is comprised of approximately 5\% baryonic matter, 26\% dark matter
and the remainder dark energy (discussed later). Some authors \citep{freese} regard the evidence of DM from the CMB as ``irrefutable''.

@@

With the advent of numerical simulations, models that involve DM have been extensively explored. Currently a very popular model of the evolution of structure in the universe is
the Lambda Cold Dark Matter ($\Lambda CDM$) model. $\Lambda$ here refers to the cosmological constant discussed in the DE section. 
Computer simulations have shown that if the DM particles are relativistic (``hot'') then, on small scales, density fluctuations are Silk damped, or
washed out, the random thermal motion of DM particles. This results in small scale structure fluctuations to be suppressed. However, if the DM particles are ``cold'' i.e. cold dark matter (CDM), then
they undergo a very different structure formation because they have a much smaller free-streaming length and can form low mass halos which can gradually build up into larger DM
structures. These DM structures provide a gravitational well that baryonic particles can fall into to form baryonic structure. This results in a bottom-up process of structure 
formation which is very different to the top-down sequence predicted for hot DM. These simulations have further convinced cosmologists that CDM comprised of WIMPs are the best
explanation for the growth of large scale structure in the universe. In fact, the $\Lambda CDM$ model agrees so well with observations that it is often regarded as the ``standard model
of cosmology''.

@@

Type Ia supernovae occur when a white dwarf in a binary pair accretes enough mass so that it goes over the Chandrasekhar mass and triggers a supernova. The supernova therefore doesn't
depend on the mass of the white dwarf on the other star in the binary. Type Ia supernovae can be considered as standard candles and enable astronomers to obtain a luminosity distance.
To measure cosmic expansion the apparent magnitude of distant supernovae are compared to those of more local supernovae which are not affected by DE. The Supernova Cosmology Project
which carried out one of the surveys in 1998 found that, with a flat cosmology ($\Omega_{\Lambda}+\Omega_m=1$), that $\Omega_m=.28$ and so that $\Omega_{\Lambda}=.72$. $\Omega$ is the
density parameter and is defined as the ratio of the observed density to the critical density. $\Omega_m$ here is 
the proportion of matter in universe (DM and baryonic) whilst $\Omega_{\Lambda}$ represents the proportion of DE in universe. The deceleration parameter is calculated as follows:
\begin{equation}
\frac{1}{2}(\Omega_m-2\Omega_{\Lambda}+2\Omega_R) 
\end{equation}
This then results in a negative deceleration parameter indicating that the expansion of the universe is accelerating.
Since 1998 there have been many more supernova surveys and it is has become an active area of observational cosmology. All this supernova data indicates that the apparent luminosity
of Type Ia supernovae declines more rapidly with increasing redshift than would be expected if $\Omega_M=1$ and $\Omega_{\Lambda}=0$ i.e. an Einstein-de Sitter universe, or the 
luminosity distance is greater than expected. Of course, these data could be interpreted as being due to other effects such as dust or gas in the line of sight. More evidence is required.

@@

There are many theories for DE. However, the simplest one is that it is constant i.e. a cosmological constant. The cosmological constant has a negative pressure equal to
its energy density and so exerts a repulsive force causing the expansion of the universe to accelerate i.e. $P=-\rho$. The previously mentioned $\Lambda CDM$ model uses
this form of DE and is the currently preferred cosmological model. An important parameter in the study of DE is the equation of state parameter $\omega$. This comes from: 
\begin{equation}
P=\omega\rho c^2
\end{equation}
where P is pressure, $\rho$ is density, c is speed of light and $\omega$ is the equation of state parameter.
In the $\Lambda CDM$ model $\omega$ is assumed to be -1.

@@

A new and exciting development by the European Space Agency, which will further probe DE, will be the launch of Euclid which is expected to be in 2020. Euclid aims to find out
the nature the acceleration of the expansion of the universe and DE. It will measure shapes and redshifts of galaxies and also explore the large-scale structure of the universe
by measuring the distribution of galaxy clusters. Euclid is a satellite equipped with a 1.2m telescope and three imaging and spectroscopic instruments working in optical and
near-infrared. It will be placed at the L2 Langrange point, a stable point in space, and is planned to operate for six years. The plan is cover 15000 square degrees of sky and
image one billion galaxies and take redshifts measurements of 100,000 galaxies \citep{amendola}. With these data Euclid will able to detail clustering of galaxies out to redshift
z=2 and out to redshift z=3 utilising weak lensing. Redshift measurements and weak lensing will also be accompanied by measuring correlations with the CMB, obtaining galaxy cluster 
information, strong lensing and possibly even obtaining luminosity distance through supernovae Type Ia measurements.

DE may well have a profound effect on the final fate of the universe. Observations suggest that the expansion of the universe will continue to accelerate forever and so the 
universe will cool as it expands. Redshift will eventually stretch photons to undetectable wavelengths and galaxies will therefore appear to disappear. Star formation will cease 
as gas supply runs out. Stars will
eventually die out, one by one, leaving stellar remnants. Some theories even predict that stellar remnants will disappear due to proton decay leaving only black holes,
which themselves will eventually disappear due to Hawking radiation. Proton decay is a hypothetical form of radioactive decay in which protons decay into lighter particles. 
Hawking radiation is blackbody radiation believed to be emitted by black holes because of quantum
effects near the event horizon. This is often referred to as the heat death of the universe. It would seem that the ultimate fate of our universe could potentially be a very 
cold, dark void.
